---
title: "Loops_and_Functions_Note_SK"
author: "Shakiba Kazemian"
date: "2025-03-27"
output: html_document
---
This module explores **iteration** and **function creation** in R using a real dataset on fungicide sensitivity. These tools help streamline analysis, reduce repetition, and make data workflows reproducible.

## Required Packages
```{r}
library(ggplot2)
library(drc)
library(tidyverse)
library(dplyr)
```


# FUNCTIONS
Functions allow us to automate repetitive tasks and write cleaner, reusable code.

## Operating with formula!! (good but not efficient)
```{r}
(5*(32 - 32)/9)  # Result: 0°C
(5*(80 - 32)/9)  # Result: ~26.7°C

```

## Instead of repeating the formula, we can write a function:

### Anatomy of a Function
```{r}
function_name <- function(input_var) {
  # some operations
  return(output)
}
```

Now, let's try writing a function that converts Celsius to Fahrenheit:
```{r}
C_to_F <- function(celsius_temp) {
  fahrenheit <- celsius_temp * (9/5) + 32
  return(fahrenheit)
}
```


# ITERATION in R
Iteration lets you repeat tasks efficiently. R offers multiple tools for this:

## rep() — Repeats elements
```{r}
rep("A", 3)
rep(c("A", "B"), 5)
rep(c(1,2,3,4), times = 4)
rep(c(1,2,5,2), times = 4, each = 4)
```

## seq() & seq_along() — Generate sequences
```{r}
seq(1, 7)
seq(0, 10, by = 2)
rep(seq(0, 10, by = 2), times = 3, each = 2)

LETTERS
seq_along(LETTERS[1:5])  # gives 1 to 5
```


# The for Loop
For loops repeat a task for each item in a sequence. Classic and widely used.

Example: Multiply numbers 1–10 by 2
```{r}
for (i in 1:10) {
  print(i * 2)
}
```


# Modeling Fungicide Sensitivity with for Loop
Goal: estimate EC50 (concentration that reduces fungal growth by 50%) for multiple isolates.

## Load and inspect the data
```{r}
EC50.data <- read.csv("EC50_all.csv")
```

## Fit model for a single isolate (ILSO_5-41c)
```{r}
isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == "ILSO_5-41c"] ~ 
        EC50.data$conc[EC50.data$is == "ILSO_5-41c"], 
        fct = LL.4(fixed = c(NA, NA, NA, NA),
                   names = c("Slope", "Lower", "Upper", "EC50")),
        na.action = na.omit)

summary.fit <- data.frame(summary(isolate1)[[3]])
EC50 <- ED(isolate1, respLev = c(50), type = "relative", interval = "delta")[[1]]
```

## Automate with a for loop

### 1. Extract unique isolate names
```{r}
nm <- unique(EC50.data$is)
```

### 2. Loop through isolates and fit model
```{r}
for (i in seq_along(nm)) {
  isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == nm[[i]]] ~ 
                  EC50.data$conc[EC50.data$is == nm[[i]]],
                  fct = LL.4(fixed = c(NA, NA, NA, NA),
                             names = c("Slope", "Lower", "Upper", "EC50")),
                  na.action = na.omit)
  
  summary.fit <- data.frame(summary(isolate1)[[3]])
  EC50 <- ED(isolate1, respLev = c(50), type = "relative", interval = "delta")[[1]]
  EC50
}
```

### 3. Save EC50 values to a dataframe
```{r}
EC50.ll4 <- NULL
for (i in seq_along(nm)) {
  isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == nm[[i]]] ~ 
                  EC50.data$conc[EC50.data$is == nm[[i]]],
                  fct = LL.4(fixed = c(NA, NA, NA, NA),
                             names = c("Slope", "Lower", "Upper", "EC50")),
                  na.action = na.omit)

  EC50 <- ED(isolate1, respLev = c(50), type = "relative", interval = "delta")[[1]]
  isolate.ec_i <- data.frame(Isolate = nm[[i]], EC50 = EC50)
  EC50.ll4 <- rbind(EC50.ll4, isolate.ec_i)
}
```

### 4. Plot EC50 distribution
```{r}
ggplot(EC50.ll4, aes(x = EC50)) +
  geom_histogram() +
  theme_classic()
```


#  map() for Functional Iteration
map() simplifies iteration using functional programming. It's concise and tidyverse-friendly.
```{r}
EC50.data %>%
  group_by(is) %>%
  nest() %>%
  mutate(ll.4.mod = map(data, ~drm(.$relgrowth ~ .$conc,
                                   fct = LL.4(fixed = c(NA, NA, NA, NA),
                                              names = c("Slope", "Lower", "Upper", "EC50"))))) %>%
  mutate(ec50 = map(ll.4.mod, ~ED(., respLev = 50, type = "relative", interval = "delta")[[1]])) %>%
  unnest(ec50)
```

