---
title: "Data_Wrangeling_ClassNotes"
author: "Shakiba Kazemian"
date: "2025-03-19"
output: html_document
---

## Data wrangeling 

The tidyverse builds upon base R to allow for easier use of large datasets. 

# loading necessary packages
```{r, warning=FALSE, echo=FALSE}
library(tidyverse)
```

# load the dataset
```{r}
microbiome.fungi <- read.csv("Bull_richness.csv")
str(microbiome.fungi)
```

#### Selecting certain columns using `select()` function
```{r}
microbiome.fungi2 <- select(microbiome.fungi, SampleID, Crop, Compartment:Fungicide, richness)
str(microbiome.fungi2)
```

#### Subsetting/filtering data using `filter()`function

This is a convenient function that allows us to subset data easily!

```{r}
head(filter(microbiome.fungi2, Treatment == "Conv."))

# A more complex using &
head(filter(microbiome.fungi2, Treatment == "Conv." & Fungicide == "C"))

# Another more complex example using or |
head(filter(microbiome.fungi2, Sample == "A" | Sample == "B")) # samples A or B
```

#### Creating a new column using `mutate()` function

```{r}
# Create a new column called logRich
head(mutate(microbiome.fungi2, logRich = log(richness)))

# Creating a new column which combines Crop and Treatment 
head(mutate(microbiome.fungi2, Crop_Treatment = paste(Crop, Treatment)))
```

#### The pipe `%>%`

Pipe allows you to input the previous data from the right side of the pipe into the function on the left (or next line) and We will combine all previous steps into one large string of functions. 

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  head() # displaying the first six rows
```

#### Summarise data using `summarise()` function

We can use the `summarise()` function to find things like means and standard deviations/errors. 
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich)) # calculating overall mean log richness within the conventionally managed treatment
```

We can also connect multiple summary statistics here. 
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```

We can also do this very easily using the `group_by()` function. 
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```

#### Connecting to plotting

It connects different data sets altogether like a package!

```{r}
plot1 <- microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n)) %>%
  ggplot(aes(x = Fungicide, y = Mean.rich)) + # adding in a ggplot
  geom_bar(stat="identity") +
  geom_errorbar( aes(x=Fungicide, ymin=Mean.rich-std.err, ymax=Mean.rich+std.err), width=0.4) +
  theme_minimal() +
  xlab("") +
  ylab("Log Richness") +
  facet_wrap(~Treatment)
```

#### Joining 

  -left_join() 
    - Keep all rows of X and add matching rows from Y. Any rows in Y that don’t match X are excluded.
  -right_join() 
    - reverse of left_join()
  -inner_join() 
    - only keep rows that are common to both X AND Y, remove everything else. 
  -full_join()
    - Keep any columns that are in either X or Y
    
Here we are also sampling 100 random rows of our dataset using the `sample_n()` function
```{r}
# selecting just the richness and sample ID
richness <- microbiome.fungi %>%
  select(SampleID, richness)

# selecting columns that don't include the richness
metadata <- microbiome.fungi %>% 
  select(SampleID, Fungicide, Crop, Compartment, GrowthStage, Treatment, Rep, Sample)

head(metadata)
head(richness)


head(left_join(metadata, richness, by = "SampleID")) # adding the richness data to the metadata based on on the common column of sampleID
```

#### Pivoting using `pivot_longer()` and `pivot_wider()` functions

Useful for converting from wide to long format and back again (mostly from wide to long).

long data: One observation per row
Wide data: Multiple observations per row

Lets explore turning our data set into a wide format, since its in long format.
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% 
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) # calculates the mean per Treatment and Fungicide 
```

Wide format - sets the values within the fungicide column into column names
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) # pivot to wide format
```

Easily can take the difference bewteen the fungicide and control now. 
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) %>% # pivot to wide format
  mutate(diff.fungicide = C - F) # calculate the difference between the means. 
```

Now we can easily calculate the difference between fungicide and control and plot it. 
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) %>% # pivot to wide format
  mutate(diff.fungicide = C - F) %>%  # calculate the difference between the means. 
  ggplot(aes(x = Treatment, y = diff.fungicide)) + # Plot it 
  geom_col() +
  theme_minimal() +
  xlab("") +
  ylab("Difference in average species richness")
```

