---
title: "Coding_Challenge4_Markdown_SK"
author: "Shakiba Kazemian"
date: "2025-02-26"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
  github_document: null
  pdf_document:
    toc: true
---

# Coding Challenge 4 - Markdown

## **Manuscript Reference**
[Noel, Z.A., Roze, L.V., Breunig, M., Trail, F. 2022. *Endophytic fungi as promising biocontrol agent to protect wheat from Fusarium graminearum head blight*. Plant Disease.](https://doi.org/10.1094/PDIS-06-21-1253-RE)

---

## **1. Explain the Following**

### a. YAML Header
A **YAML (Yet Another Markup Language) header** in R Markdown is a section at the top of the file enclosed by triple dashes (`---`). It defines metadata for the document, such as the title, author, date, and output format.

### b. Literate Programming
Literate programming means **writing programs as documents**, where explanations guide the reader, and code is embedded seamlessly. **R Markdown is a perfect example of literate programming!**

---

## **2. Data Preparation**
### **Read the Data**
The dataset is loaded using a **relative file path**, ensuring portability. Missing values (`NA`) are explicitly defined.

```{r load_data, echo=TRUE}
# Load necessary libraries
knitr::opts_knit$set(root.dir = getwd())  # Ensure consistent working directory
library(ggplot2)
library(ggpubr)
library(dplyr)

# Read the data using a relative path
data <- read.csv("R_Markdown/Coding_Challenge4/MycotoxinData.csv", na.strings = "na")

# Display the first few rows of the dataset
head(data)
```

### **Data Visualization**
#### Plotting Individual Figures
Separate plots for DON data, 15ADON, and Seedmass using ggplot2.
```{r}
# DON Plot
DON_plot <- ggplot(data, aes(x = Treatment, y = DON)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("DON Levels by Treatment")

# 15ADON Plot
ADON_plot <- ggplot(data, aes(x = Treatment, y = X15ADON)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("15ADON Levels by Treatment")

# Seedmass Plot
SeedMass_plot <- ggplot(data, aes(x = Treatment, y = MassperSeed_mg)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Seed Mass per mg by Treatment")
```

### **Statistical Analysis: Pairwise t-tests**
Adding pairwise t-tests to each plot using stat_compare_means().
```{r}
# Add Pairwise t-tests to Each Plot
DON_plot_ttest <- DON_plot + stat_compare_means(method = "t.test", aes(group = Treatment), label = "p.signif")
ADON_plot_ttest <- ADON_plot + stat_compare_means(method = "t.test", aes(group = Treatment), label = "p.signif")
SeedMass_plot_ttest <- SeedMass_plot + stat_compare_means(method = "t.test", aes(group = Treatment), label = "p.signif")
```

### **Combining All Plots**
Using ggarrange() to create a single figure with all plots.
```{r combine_plots, warning=FALSE, message=FALSE}
# Combine all updated plots into one figure
Combined_Plot_ttest <- ggarrange(
  DON_plot_ttest, ADON_plot_ttest, SeedMass_plot_ttest,  
  labels = c("A", "B", "C"),  
  ncol = 3, nrow = 1,  
  common.legend = TRUE,  
  legend = "right"
)

# Display the final combined plot with t-tests
print(Combined_Plot_ttest)

# Save the final combined figure
ggsave("Combined_Plot_ttest.png", Combined_Plot_ttest, width = 12, height = 5, dpi = 300)
```
