---
title: "CodingChallenge5_DataWrangeling_SK"
author: "Shakiba Kazemian"
date: "2025-03-20"
output: pdf_document
---

# Q1: Reading the CSV files using relative paths

```{r}
# Loading necessary package
library(tidyverse)

# Read the files
diversity_data <- read.csv("DiversityData.csv")
metadata <- read.csv("Metadata.csv")

# Check structure of the data
str(diversity_data)
str(metadata)
```


# Q2: Joining the two dataframes together by the common column ‘Code’.

```{r}
# Join the two dataframes by the common column "Code"
alpha <- left_join(metadata, diversity_data, by = "Code")

# Check the structure of the resulting dataframe
str(alpha)

# Display the first few rows
head(alpha)
```


# Q3: Calculating Pielou’s evenness index: Pielou’s evenness is an ecological parameter calculated by the Shannon diversity index (column shannon) divided by the log of the richness column. 

```{r}
# Create a new column for Pielou’s evenness index
alpha_even <- alpha %>%
  mutate(Pielou_Evenness = shannon / log(richness))

# Check the structure of the new dataframe
str(alpha_even)

# Display the first few rows
head(alpha_even)
```


# Q4: Calculating the mean and standard error evenness grouped by crop over time.

```{r}
# Summarize the data: Mean and standard error of Pielou's evenness index grouped by Crop and Time_Point
alpha_average <- alpha_even %>%
  group_by(Crop, Time_Point) %>%  # Group by Crop and Time_Point
  summarise(
    Mean_Evenness = mean(Pielou_Evenness, na.rm = TRUE),  # Calculate mean evenness
    n = n(),  # Count the number of observations per group
    sd_evenness = sd(Pielou_Evenness, na.rm = TRUE)  # Standard deviation
  ) %>%
  mutate(std_err_evenness = sd_evenness / sqrt(n))  # Calculate standard error

# Check the structure of the resulting dataframe
str(alpha_average)

# Display the first few rows
head(alpha_average)

```


# Q5: Calculating the differences in mean evenness between Soybean, Cotton, and Soil.

```{r}
# Transform the data from long to wide format
alpha_average2 <- alpha_average %>%
  select(Time_Point, Crop, Mean_Evenness) %>%  # Select relevant columns
  pivot_wider(names_from = Crop, values_from = Mean_Evenness) %>%  # Reshape to wide format
  mutate(
    diff.cotton.even = Soil - Cotton,   # Difference between Cotton and Soil
    diff.soybean.even = Soil - Soybean   # Difference between Soybean and Soil
  )

# Check the structure of the resulting dataframe
str(alpha_average2)

# Display the first few rows
head(alpha_average2)
```


# Q6: creating the plot using ggplot2 after reshaping the data with pivot_longer()

```{r}
# Reshape the data from wide to long format
alpha_plot_data <- alpha_average2 %>%
  select(Time_Point, diff.cotton.even, diff.soybean.even) %>%  # Select relevant columns
  pivot_longer(cols = c(diff.cotton.even, diff.soybean.even), 
               names_to = "diff", 
               values_to = "values")  # Reshape data

# Create the plot
ggplot(alpha_plot_data, aes(x = Time_Point, y = values, color = diff, group = diff)) +
  geom_line(size = 1) +  # Line plot
  geom_point(size = 2) +  # Add points
  theme_minimal() +  # Minimal theme
  labs(
    x = "Time (hrs)",  # X-axis label
    y = "Difference from soil in Pielou's evenness",  # Y-axis label
    color = "diff"  # Legend title
  ) +
  scale_color_manual(values = c("diff.cotton.even" = "salmon", "diff.soybean.even" = "skyblue")) +  # Adjust colors
  theme(
    text = element_text(size = 12),  # General text size
    axis.title.x = element_text(size = 14),  # X-axis title size
    axis.title.y = element_text(size = 14),  # Y-axis title size
    axis.text = element_text(size = 12),  # Tick labels size
    legend.text = element_text(size = 12),  # Legend text size
    legend.title = element_text(size = 12)  # Legend title size
  )
```


# Q7: This is my submission for Coding Challenge 5.

[Click here to view my GitHub repository](https://github.com/szk0263/PLPA6820_SK/tree/main/Data_Wrangeling/Coding_Challenge5)

