---
title: "CodingChallenge7_LinearModel_SK"
author: "Shakiba Kazemian"
date: "2025-04-03"
output:
  pdf_document: default
  md_document:
    variant: markdown_github
---

## Q1: Reading the data
```{r}
# Load required libraries
library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
library(multcompView)

# Read in the data
PlantEmergence <- read.csv("PlantEmergence.csv")

# Convert columns to factors
PlantEmergence$Treatment <- as.factor(PlantEmergence$Treatment)
PlantEmergence$DaysAfterPlanting <- as.factor(PlantEmergence$DaysAfterPlanting)
PlantEmergence$Rep <- as.factor(PlantEmergence$Rep)
```

## Q2: Fitting the linear model with Treatment, DaysAfterPlanting, and their interaction.
```{r}
# Fit the linear model with interaction
lm_emergence <- lm(Emergence ~ Treatment * DaysAfterPlanting, data = PlantEmergence)

# View the summary of the linear model
summary(lm_emergence)

# View the ANOVA table
anova(lm_emergence)
```

## Q3: 
Based on the ANOVA results, the interaction between Treatment and DaysAfterPlanting is not significant (p = 1.000), so we do not need to include it in the model. A simplified model with only the main effects of Treatment and DaysAfterPlanting is sufficient.

### Step1: fit simplified linear model
```{r}
# Fit simplified linear model with only main effects
lm_simple <- lm(Emergence ~ Treatment + DaysAfterPlanting, data = PlantEmergence)

# Summary of the linear model
summary(lm_simple)

# ANOVA table for the model
anova(lm_simple)

```

### Step2: Interpretation
The simplified model with excluding the interaction term fits the data very well (Adjusted R² = 0.9545).
Both Treatment and DaysAfterPlanting are significant predictors of Emergence.
The intercept (182.16) represents average emergence for Treatment 1 at baseline planting day.
The Treatment2 coefficient (-134.53) indicates that, all else equal, Treatment 2 results in a massive and statistically significant reduction in emergence.

## Q4: calculating least square means (LS means) for Treatment, performing Tukey’s post-hoc test, and interpreting the Compact Letter Display (CLD) using the emmeans and cld() functions.
```{r}
# Load emmeans and multcompView if not already loaded
library(emmeans)
library(multcompView)

# Calculate LS means (estimated marginal means) for Treatment
treatment_lsmeans <- emmeans(lm_simple, ~ Treatment)

# Tukey post-hoc test with compact letter display
treatment_cld <- cld(treatment_lsmeans, alpha = 0.05, Letters = letters, reversed = TRUE)

# View results
treatment_cld
```
### Conclusion:
Treatments 3, 5, and 6 had the highest and statistically similar emergence.
Treatment 2 is significantly worse than all others and should likely be avoided.
Intermediate treatments (like 4, 8, 9, 1) may perform acceptably but are not clearly top-tier.

## Q5: Making a plot using the provided function

### Step1: Running the full function definition
```{r}
plot_cldbars_onefactor <- function(lm_model, factor) {
  data <- lm_model$model
  variables <- colnames(lm_model$model)
  dependent_var <- variables[1]
  independent_var <- variables[2:length(variables)]

  lsmeans <- emmeans(lm_model, as.formula(paste("~", factor))) # estimate lsmeans 
  Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE, Letters = letters) # contrast with Tukey adjustment by default.

  # Extracting the letters for the bars
  sig.diff.letters <- data.frame(Results_lsmeans$emmeans[,1], 
                                 str_trim(Results_lsmeans$emmeans[,7]))
  colnames(sig.diff.letters) <- c(factor, "Letters")

  # for plotting with letters from significance test
  ave_stand2 <- lm_model$model %>%
    group_by(!!sym(factor)) %>%
    dplyr::summarize(
      ave.emerge = mean(.data[[dependent_var]], na.rm = TRUE),
      se = sd(.data[[dependent_var]]) / sqrt(n())
    ) %>%
    left_join(sig.diff.letters, by = factor) %>%
    mutate(letter_position = ave.emerge + 10 * se)

  plot <- ggplot(data, aes(x = !! sym(factor), y = !! sym(dependent_var))) + 
    stat_summary(fun = mean, geom = "bar") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
    ylab("Number of emerged plants") + 
    geom_jitter(width = 0.02, alpha = 0.5) +
    geom_text(data = ave_stand2, aes(label = Letters, y = letter_position), size = 5) +
    xlab(as.character(factor)) +
    theme_classic()

  return(plot)
}

```

### Make the plot
```{r}
plot_cldbars_onefactor(lm_simple, "Treatment")
```

## Link to my GitHub

[Click here to view my submission on GitHub](https://github.com/szk0263/PLPA6820_SK/tree/main/Linear_Models/Coding_Challenge7)
